Active Type Constructors

### 2014-07-22 atlang-gpce14/atlang-gpce14.tex
GPCE14 post-mortem

\lefthyphenmin=62
\sloppy
%\makeatletter
%\renewcommand\section{\@startsection{section}{1}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\large\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsection{\@startsection{subsection}{2}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\normalsize\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
%                       {-5\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\bfseries\boldmath}}
%\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
%                       {-4\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\itshape}}
%\makeatother

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers justifiably prefer abstractions available as libraries for an existing language over those requiring a switch to a new language, but library providers cannot extend these language's type systems directly, so powerful statically typed abstractions often remain obscure. This suggests a need for a language that is \emph{backwards compatible} with the libraries, tools and infrastructure of a widely adopted language but that has an  \emph{internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported  without  ambiguity.

Our aim in this paper is thus to 1) introduce such a mechanism and 2) show how it can be implemented in a backwards compatible manner within an existing widely used language. We do so with two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term constructor by delegating control, according to a novel bidirectional typechecking and translation protocol, to methods associated with a relevant \emph{active type constructor}, defined in a library. The type constructor delegated control over a term is stable even when additional extensions are imported, so we avoid ambiguities by construction. % By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a library is simpler than asking them to adopt a new programming language. Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.

Unfortunately, researchers and domain experts designing new abstractions can  find it difficult to soundly and naturally implement them as libraries, particularly when these require strengthening the language's static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in a foreign language requires programming against (and thus revealing) its internal implementation. This is brittle, unnatural and unsafe and can cause performance issues due to additional necessary checks and conversions at language boundaries. %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

@@

\inferrule[att-ana-lam]{
    \sigma_1~\mathtt{ty}_\fvalCtx\\\sigma_2~\mathtt{ty}_\fvalCtx\\
  \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e \Leftarrow \sigma_2 \leadsto \iota
}{
  \atjanaX{\lambda x.e}{\FF{ty}[\fvar{fn}](\FF{cons}(\sigma_1; \FF{cons}(\sigma_2; \FF{nil})))}{\lambda(x.\iota)}
}

\inferrule[att-syn-idx-lam]{
    \sigma_1~\mathtt{ty}_\fvalCtx\\
  \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e \Rightarrow \sigma_2 \leadsto \iota
}{
  \atjsynX{\FF{asc}[\FF{incty}[\fvar{fn}](\sigma_1)](\lambda(x.e))}{\FF{ty}[\fvar{fn}](\FF{cons}(\sigma_1; \FF{cons}(\sigma_2; \FF{nil})))}{\lambda(x.\iota)}
}
%
%\inferrule[att-lam]{
% \tau_1 \Downarrow_{\Gamma;\fvalCtx} \tau_1'\\
%   \trepof{\tau_1'} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\\\
% \Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto \iota
%%  \iota \hookrightarrow_\fvalCtx \iota'\\
%%    \sigma \hookrightarrow_\fvalCtx \sigma'
%%  \ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
% %\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
%}{
% \atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{\iota}}
%}

>>>>

\lefthyphenmin=8
\sloppy
%\makeatletter
%\renewcommand\section{\@startsection{section}{1}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\large\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsection{\@startsection{subsection}{2}{\z@}%
%                       {-18\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {8\p@ \@plus 4\p@ \@minus 4\p@}%
%                       {\normalfont\normalsize\bfseries\boldmath
%                        \rightskip=\z@ \@plus 8em\pretolerance=10000 }}
%\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
%                       {-5\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\bfseries\boldmath}}
%\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
%                       {-4\p@ \@plus -4\p@ \@minus -4\p@}%
%                       {-0.5em \@plus -0.22em \@minus -0.1em}%
%                       {\normalfont\normalsize\itshape}}
%\makeatother

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers justifiably prefer abstractions available as libraries for the language they already use over those requiring a switch to a new language. But library providers cannot typically extend the type systems of these languages directly, so powerful statically typed abstractions  remain obscure. This suggests that a language is needed that is \emph{backwards compatible} with the libraries and tools of a widely adopted language, but that has an  \emph{internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported  without  ambiguity.

Our aim in this paper is to 1) introduce such a mechanism, and 2) show how it can itself be implemented as a library for an existing widely used language. We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal core calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term constructor by delegating control, according to a novel bidirectional typechecking and translation protocol, to methods associated with a relevant \emph{active type constructor}, defined in a library. The delegate is stable even when additional extensions are imported, so we avoid ambiguities by construction. We demonstrate with several examples of interesting type system fragments as libraries.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a library is easier than asking them to adopt a new programming language. Indeed, recent empirical studies underscore the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with large existing code bases, library availability, familiarity and tool support are at least as important as intrinsic features of the language \cite{Meyerovich:2013:EAP:2509136.2509515,chen05}.

Unfortunately, researchers and domain experts designing new abstractions can  find it difficult to soundly and naturally implement them as libraries, particularly when they require strengthening the language's static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in a foreign language requires programming against (and thus revealing) its internal implementation. This is brittle, unnatural and unsafe and can cause performance issues due to additional necessary checks and conversions at language boundaries. %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

@@

\inferrule[att-ana-lam]{
   \Phi_\text{fn} \subset \Phi\\ \sigma_1~\mathtt{ty}_\fvalCtx\\\sigma_2~\mathtt{ty}_\fvalCtx\\
  \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e \Leftarrow \sigma_2 \leadsto \iota
}{
  \atjanaX{\lambda(x.e)}{\FF{ty}[\fvar{fn}](\FF{cons}(\sigma_1; \FF{cons}(\sigma_2; \FF{nil})))}{\lambda(x.\iota)}
}

\inferrule[att-syn-idx-lam]{
   \Phi_\text{fn} \subset \Phi\\ \sigma_1~\mathtt{ty}_\fvalCtx\\
  \Gamma, x \Rightarrow \sigma_1 \vdash_\fvalCtx e \Rightarrow \sigma_2 \leadsto \iota
}{
  \atjsynX{\FF{asc}[\FF{incty}[\fvar{fn}](\sigma_1)](\lambda(x.e))}{\FF{ty}[\fvar{fn}](\FF{cons}(\sigma_1; \FF{cons}(\sigma_2; \FF{nil})))}{\lambda(x.\iota)}
}
%
%\inferrule[att-lam]{
% \tau_1 \Downarrow_{\Gamma;\fvalCtx} \tau_1'\\
%   \trepof{\tau_1'} \Downarrow_{\Gamma;\fvalCtx} \titype{\sigma}\\\\
% \Gamma, x \Rightarrow \tau_1' \vdash_\fvalCtx e \Rightarrow \tau_2 \leadsto \iota
%%  \iota \hookrightarrow_\fvalCtx \iota'\\
%%    \sigma \hookrightarrow_\fvalCtx \sigma'
%%  \ddbar{\fvar{Arrow}}{\fvalCtx}{\trepof{\tau_1'}}{\sbar_1}\\
% %\delfromtau{$\Xi_0$}{\fvalCtx}{\tau_1'}{\sabs}\\\\
%}{
% \atjsynX{\elam{x}{\tau_1}{e}}{\ttype{arrow}{(\tau_1', \tau_2)}}{\ilam{x}{\sigma'}{\iota}}
%}
